Robot Team III: Scotty Strikes Back

### 2024-04-30 paper/paper.tex
moved things from the draft intro to the main document

Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability \cite{koponen2002lake,ritchie2003remote}. Hyperspectral imaging has emerged as a pivotal technology in this space, where the heightened spectral resolution enables the inversion of optically-active water quality parameters such as colored dissolved organic matter (CDOM), chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{brando2003satellite,moses2012estimation}.

Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches to map observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral,keller2018hyperspectral, ghatkar2019classification,sagan2020monitoring}. However, the calibration and evaluation of these models demands a significant volume of coincident, in situ data. Collecting paired datasets is time consuming and expensive, often relying on rare satellite overpasses spanning decades \cite{aurin2018remote, ross2019aquasat}. This requirement impedes the real-time assessment of water quality change, for example, during oil spills \cite{fingas2014review,fingas2017review}.

@@

due to the dramatic improvement in spectral resolution afforded by HSI platforms. Specifically, the combination with hyperspectral imaging together with autonomous collection by UAVs enables unparalleled spectral, spatial, and temporal resolution. Significant effort has developed the use of hyperspectral imagery for the inversion of optically-active water quality parameters. However, this approach requires access to sufficient quantities of in situ reference data to fit and evaluate each model. Traditionally, this process has required collocation of remote sensing imagery with data from field campaigns spanning decades (cite Aurin and Ross). Recently, we have demonstrated how combining autonomous collection of in situ data with UAV-borne hyperspectral imagery dramatically reduces the time required to collect these data. Both of these approaches require we already know what to look for.

Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources.

@@

@@

>>>>

Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability \cite{koponen2002lake,ritchie2003remote}. Hyperspectral imaging has emerged as a pivotal technology in this space, where the \textcolor{purple}{heightened} spectral resolution of hyperspectral images (HSIs) enables the inversion of optically-active water quality parameters such as colored dissolved organic matter, chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{brando2003satellite,moses2012estimation}. Machine learning techniques including tree-based models, support vector machines, deep neural networks, and other architectures have proven effective at mapping observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral,keller2018hyperspectral, ghatkar2019classification,sagan2020monitoring}. However, the calibration and evaluation of these models demands a significant volume of coincident, in situ data. Collecting these paired datasets is time consuming and expensive, often relying on rare satellite overpasses spanning decades of collection \cite{aurin2018remote, ross2019aquasat}. This requirement impedes the real-time assessment of water quality change, for example, during oil spills \cite{fingas2014review,fingas2017review}.

@@

The continued development of hyperspectral imaging technology has resulted in significant size reductions enabling the deployment of hyperspectral imagers on unmanned aerial vehicles (UAVs) such as drones. UAVs are not only less expensive than their remote sensing counterparts but also limit the need for complicated atmospheric corrections and enable centimeter scale spatial resolution through low altitude flights \cite{adao2017hyperspectral, arroyo2019implementation, banerjee2020uav}.

- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}

UAV-based hyperspectral imagers have been used estimate chlorophyll-a, suspended solids, and  
Many such systems have been utilized for the characterization of water quality and composition (cite examples).
\cite{lu2021retrieval}
\cite{zhang2022selection}

The collection of requisite in situ data can be significantly accelerated by coordinating UAV flights with reference-sensor equipped autonomous boats \cite{robot-team-1, robot-team-2}. 
However, this approach requires prior knowledge of expected contaminants in order to select relevant sensors

- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}

Endmember extraction and unmixing approaches can leverage the ability of hyperspectral images to distinguish source types. Example: mineral identification from AVRIS data, algae classification for harmful algal bloom detection, and distinguishing types of oils based on slight difference in spectral shapes.

@@

Unsupervised classification
- When no ground-truth data are available, unsupervised classification can still help partition imagery into groups or clusters.
- Many data-driven, ML methods have been employed for the task including various matrix factorizations, k-nearest neighbors, fuzzy c-means, density estimation methods, etc. \cite{zhang2019hyperspectral}
- Additionally unsupervised approaches can be used to perform nonlinear dimensionality reduction & pre-processing for supervised approaches.
- Many unsupervised approaches can be challenging to interpret due to the lack of a relationship between the identified classes, e.g. is class 1 more similar to class 2 than class 10?
- The Self Organizing Map is a popular technique which classifies data while enforcing topological relationships between the resulting classes such that classes close to each  other in the latent space are more similar than classes that are far apart.
- SOM used for remote sensing imagery classification \cite{msom-remote-sensing}
- SOM used for clustering and data compression of HSI cube-sat \cite{danielsen2021self}
- SOM used for land-use and land-cover change analysis \cite{penfound2021analysis}
- SOM has been used for identifying synoptic-scale patterns in wind and sea surface temperature data {som-satellite}

endmember extraction, spectral unmixing
- This is a related problem where the goal is to identify unique spectral signatures which combine (linearly or non-linearly) to produce the measured signal.
- Having identified the set of endmembers, we then seek to determine their relative abundance in each pixel
- Many statistical approaches for extracting endmembers \cite{berman2004ice}
- Sparse PCA used for endmember extraction \cite{yousefi2016mineral}
- Popular spectral unmixing approach is to treat "pure" endmembers as vertices of a simplex \cite{plaza2012endmember, nascimento2005vertex}
- Convolutional Neural Networks with Autoencoder architectures are a popular ML approach which identify endmembers and perform non-linear unmixing \cite{palsson2020convolutional, non-negative-autoencoders, su2019daen, borsoi2019deep}
- Self Organizing Map is another approach which has been used for endmember extraction together with neural networks for abundance mapping (unmixing) \cite{cantero2004analysis}

@@

    - The generative topographic mapping is an attractive algorithm 
        - Probabilistic interpretation
        - Topographic relationship maintained between classes (not just orthogonality)
        - clear fitting criterion with Maximum Likelihood Estimation / Maximum A-Posteriori Estimation
    - We use the GTM to analyze hyperspectral imagery acquired by a drone over pond to 
        - Present method for unsupervised characterization of small-scale variability of pond (i.e. identifying regions of interest which can be used to direct robot boat for further instructions) 
        - Use GTM to extract endmember for algae and map the algal distribution near shore (relevant for Harmful Algal Bloom detection)
        - Use GTM to extract endmember for rhodamine dye and use it to mape evolution of dye plume. This is a demonstration of source identification and mapping. Relevant for the identification of pollution events such as oil spills, industrial seepage, etc...

To support and enhance real-time decision making, data-driven techniques...

Using data collected at site in Montague, North Texas, we demonstrate the ability of this paradigm with two case studies. First, we demonstrate how the GTM can be used to provide a high-resolution unsupervised classification of a water body enabling the rapid identification of relevant regions which can then be used to direct further investigation by reference-sensor equipped boats. Second, we demonstrate the ability to perform endmember extraction and abundance mapping using imagery from a test-release of rhodamine dye. Together with the normalized spectral similarity score, these spectral signatures are identified and their associated area in the scene is determined.
